Cache Oblivious B-trees

### 2011-12-09 paper.tex
Adding the cituations

Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
\cite{cobtree}. The CO B-tree achieves its number of memory transfers by structuring its elements as a
strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
a packed memory array (PMA) using a van Emde Boas (vEB) layout. The SWBST and
PMA play crucial roles in the other cache-oblivious data structures that were
developed after the CO B-tree and thus are described at length in the next two
sections.

@@

The ultimate goal of a CO B-tree is to reduce the number of memory transfers
while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, at a
fundamental level the CO B-tree still needs to maintain its data in some
structure. That structure is a balanced search tree. The search tree is laid
out according to a van Emde Boas (vEB) layout [???, ???]. The idea of a vEB
layout is fairly straight forward. Let $h$ denote the height of a tree. the
vEB layout splits the tree approximately at height $h/2$. This splits the
original tree into a ``top'' subtree and many ``bottom'' subtrees. Let $A$
denote the top subtree and $B_1... B_k$ denote the bottom subtrees. The vEB
layout is defined recursively using the order: $A, B1, ... B_k$. This layout
is also shown in Figure~\ref{fig:veb}.

A nice consequence of using a vEB layout is that it is known that it will use
$O(log_B N)$ memory transfers for searches [???, ???]. Thus, the CO B-tree has already
achieved the desired number of memory transfers for search queries. It
still remains to be shown that the CO B-tree achieves the desired number of
memory transfers for dynamic inserts and deletes.

@@

A known data structure that achieves this property is the ``weight-balanced
B-tree'' \cite{????}. The CO B-tree leverages this data structure to maintain
its elements. Before SWBST can be used, however, it remains to be proven that
it maintains the the $c^h$ descendants property even after inserts and
deletes.

The definition of a SWBST given in [????] provides us with this bound for the
weight of a node:

@@

Now that we have our elements structured in a SWBST, we are still left with
mapping this tree to memory and proving that it achieves our desired number of
memory transfers. The packed memory array (PMA) was developed to solve this
problem [???]. There are three main costs that the PMA is concerned with.

Fist, the PMA needs to maintain open space in its memory for newly inserted
items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
transfers. Doing so may seem like attempting to achieve contradictory goals:
maintaining the locality of elements while keeping space available for new
inserts. The PMA uses a technique described in [???]. At a high level, the PMA
maintains a windows into the array. When a window becomes too full or too
empty, the elements are spread out within the context of a larger window. Two
parameters are required, the window size and the threshold for trigger the
spreading out. However, these parameters are not related to the block size of
the memory level and thus do not violate the CO model of computation. Overall,
the PMA is able to keep extra space in $O(1 + (log^2 N) / B)$ memory
transfers.

Second, the PMA needs to handle the splitting and merging that will be caused
by the SWBST discussed in the previous section. The PMA is able to accomplish
this using $O(1 + (log N) /B)$ memory transfers. The proof is based on the
fact that each node has $c^h$ descendants (recall, this is the key property
that SWBST provided). Thus, merging or splitting a node will cause $c^h$ nodes
to move, requiring $O(1 + c^h/B)$ memory transfers. It follows from the
definition of SWBST in [???] that when a node is rebalanced, it will require
$c^h$ insertions or deletions until it is rebalanced again. Thus, the
amortized cost of each insertion and/or deletion is $O(1/B)$. When you split a
node, it may go all the way up the tree and thus overall it will incur $O((log
N) / B)$ memory transfers.

@@

Therefore, overall, the CO B-Tree achieve insertions and deletions in $O(log_B N +
(log^2N)/B)$ memory transfers. However, it still achieves the optimal
$O(log_B N)$ memory transfers for searches, matching the performance of
ordinary, cache-aware, B-trees.

>>>>

Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
\cite{BenderDemainColton}. The CO B-tree achieves its number of memory transfers by structuring its elements as a
strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
a packed memory array (PMA) using a van Emde Boas (vEB) layout. The SWBST and
PMA play crucial roles in the other cache-oblivious data structures that were
developed after the CO B-tree and thus are described at length in the next two
sections.

@@

The ultimate goal of a CO B-tree is to reduce the number of memory transfers
while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, at a
fundamental level the CO B-tree still needs to maintain its data in some
structure. That structure is a balanced search tree. The search tree is laid
out according to a van Emde Boas (vEB) layout \cite{veb1}, \cite{veb2}. The idea of a vEB
layout is fairly straight forward. Let $h$ denote the height of a tree. the
vEB layout splits the tree approximately at height $h/2$. This splits the
original tree into a ``top'' subtree and many ``bottom'' subtrees. Let $A$
denote the top subtree and $B_1... B_k$ denote the bottom subtrees. The vEB
layout is defined recursively using the order: $A, B1, ... B_k$. This layout
is also shown in Figure~\ref{fig:veb}.

A nice consequence of using a vEB layout is that it is known that it will use
$O(log_B N)$ memory transfers for searches \cite{veb1}, \cite{veb2}. Thus, the CO B-tree has already
achieved the desired number of memory transfers for search queries. It
still remains to be shown that the CO B-tree achieves the desired number of
memory transfers for dynamic inserts and deletes.

@@

A known data structure that achieves this property is the ``weight-balanced
B-tree'' \cite{swbst}. The CO B-tree leverages this data structure to maintain
its elements. Before SWBST can be used, however, it remains to be proven that
it maintains the the $c^h$ descendants property even after inserts and
deletes.

The definition of a SWBST given in \cite{swbst} provides us with this bound for the
weight of a node:

@@

Now that we have our elements structured in a SWBST, we are still left with
mapping this tree to memory and proving that it achieves our desired number of
memory transfers. The packed memory array (PMA) was developed to solve this
problem \cite{BenderDemainColton}. There are three main costs that the PMA is concerned with.

Fist, the PMA needs to maintain open space in its memory for newly inserted
items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
transfers. Doing so may seem like attempting to achieve contradictory goals:
maintaining the locality of elements while keeping space available for new
inserts. The PMA uses a technique described in \cite{pma}. At a high level, the PMA
maintains a windows into the array. When a window becomes too full or too
empty, the elements are spread out within the context of a larger window. Two
parameters are required, the window size and the threshold for trigger the
spreading out. However, these parameters are not related to the block size of
the memory level and thus do not violate the CO model of computation. Overall,
the PMA is able to keep extra space in $O(1 + (log^2 N) / B)$ memory
transfers.

Second, the PMA needs to handle the splitting and merging that will be caused
by the SWBST discussed in the previous section. The PMA is able to accomplish
this using $O(1 + (log N) /B)$ memory transfers. The proof is based on the
fact that each node has $c^h$ descendants (recall, this is the key property
that SWBST provided). Thus, merging or splitting a node will cause $c^h$ nodes
to move, requiring $O(1 + c^h/B)$ memory transfers. It follows from the
definition of SWBST in \cite{swbst} that when a node is rebalanced, it will require
$c^h$ insertions or deletions until it is rebalanced again. Thus, the
amortized cost of each insertion and/or deletion is $O(1/B)$. When you split a
node, it may go all the way up the tree and thus overall it will incur $O((log
N) / B)$ memory transfers.

@@

Therefore, overall, the CO B-Tree achieves insertions and deletions in 
$O(log_B N + (log^2N)/B)$ memory transfers. However, it still achieves the optimal
$O(log_B N)$ memory transfers for searches, matching the performance of
ordinary, cache-aware, B-trees.
